Leadership across cultures and contexts: an exploratory analysis

### 2020-05-12 leadership_across_cultures_contexts.Rmd
Minor edits in discussion.

These potentially universal dimensions of leadership correspond to those highlighted by several evolutionary theories of leadership that have drawn heavily on select ethnographic cases [@garfield_review_2019]. They partially correspond to the prestige model [@henrich_big_2015; @henrich_evolution_2001], for example, which emphasizes knowledge, skill and generosity, but they do not include a key feature of that model: emulation of prestigious leaders. We labeled one of our leader quality features "cultural conformity", however, in which leaders are culturally conservative and exemplify desirable traits such as honesty, humility, and "proper behavior" (Figure \@ref(fig:cluster)). This feature might correspond, at least in part, to leaders who serve as models for their followers [see also @garfield_investigating_2019].

@@

Most potentially universal dimensions discovered here appear in a model developed by @hagen_leadership_2019, in which knowledgeable and intelligent individuals provide computational (cognitive) services to group members in exchange for a variety of fitness benefits, and these services often involve conflict resolution and other social functions [see also @garfield_evolutionary_2019]. The computational services model emphasizes mating benefits, however, which appear in `r matingbenefit`% of cultures, below our arbitrary 60% threshold for "universality" (see Figure \@ref(fig:varsupportcb)). But there was evidence for the "Social, reproductive, and materials success" feature, which included leaders in polygynous unions, having a high quality spouse, and having many children, in more cultures (`r reprosuccess`%) than any other feature (see Figure \@ref(fig:featuressupportplot)). Interestingly, the computational model highlights women's family leadership role, and the estimated marginal mean for the *Social, reproductive, and materials success* feature was higher for females leaders than male leaders (see Figure \@ref(fig:heatmapfeaturesex)), with the caveat that there were few text records on female leaders in our sample.

@@

These factors were confounded, i.e., group context was associated with subsistence strategy, which was associated with continental region. See Figures \@ref(fig:grouptype) and \@ref(fig:regionsubsistence). Specifically, evidence for leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013; @venkataraman_hunter-gatherer_2017]. Evidence for supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories. These patterns support perspectives suggesting residential mobility and increased costs of dispersal, subsistence intensification, and demographic transitions are associated with increasing complexity of sociopolitical organization among non-WEIRD populations [@johnson_evolution_1987; @steward_basin-plateau_1938; @powers_evolutionary_2014; @kantner_identifying_2010]. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese by @az02-001[p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in Methods section).

@@

Leaders were widely reported to receive material benefits, social status, and mating benefits but loss of material resources and social status were also widely reported, suggesting leadership to be a high-risk, high-return strategy (see Figure \@ref(fig:varsupportcb)). The results were similar for followers: followers benefited from increased material and social resources, but also often lost material and social resources. Followers were not reported to receive mating benefits, relative to leaders or other types benefits. Our results cannot determine if the benefits to leaders and followers outweighed their costs, but they do shed light on the social and biological pressures associated with leadership and social hierarchy. For both leaders and followers, individuals stood to lose what they were able to gain. See Figure \@ref(fig:varsupportcb).

@@

**Dominance and prestige**. Much of the evolutionary theoretical literature on leadership has suggested social influence is most often achieved via one of two distinct strategies: dominance or prestige [for review see @garfield_review_2019; @garfield_evolutionary_2019]. There is a long history of leadership styles being described as based on either force and strength (e.g., dominance) or respect and special skills (e.g., prestige). Such dual-models of leadership styles are rooted in the coupling of ethology and ethnography [see @mead_sex_1935; @sahlins_poor_1963; @lewis_leaders_1974; @tiger_dominance_1970; @bernstein_dominance_1976]. Synthesizing the ethnographic and ethological records, @kracke_force_1978 linked dominance and prestige to psychological-affect between leaders and followers in small-scale societies and @barkow_prestige_1980 suggested the evolution of prestige was linked to sexual selection and symbolic culture. The dominance-prestige theory as formulated by @henrich_evolution_2001 significantly advanced evolutionary approaches to leadership and social hierarchy and inspired much subsequent work across the behavioral sciences [e.g., @van_vugt_dual_2019; @cheng_two_2013; @maner_2016; @halevy_status_2012; @suessenbach_dominance_2019; @maner_dominance_2017; @garfield_evolutionary_2019; @garfield_investigating_2019]. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on distinctions between dominance and prestige suggests, "human social stratification is characterized by the co-occurrence of both strategies." Exploratory cluster analyses reported here, however, suggest multiple features of leader qualities and leader functions, beyond dominance and prestige.

@@

<!-- DISCUSSES BOTH MANAGEMENT FEATURES -->
**Leaders are managers**. In the absence of state-institutions local leaders are often tasked with managing various economic systems and providing social services. Within the domain of leader management, there are likely to be two primary features of human leader functions: strategizing and organizing. Our results correspond very closely to what @van_vugt_cognitive_2007 define as *strategic* leadership, where leaders increase the pay-offs of another behavior thereby incentivizing followership versus *coordinating* leadership, where individuals bound by group challenges must effectively organize collective effort.

>>>>

These potentially universal dimensions of leadership correspond to those highlighted by several evolutionary theories that have drawn heavily on select ethnographic cases [@garfield_review_2019]. They partially correspond to the prestige model [@henrich_big_2015; @henrich_evolution_2001], for example, which emphasizes knowledge, skill and generosity, but they do not include a key feature of that model: emulation of prestigious leaders. We labeled one of our leader quality features "cultural conformity", however, in which leaders are culturally conservative and exemplify desirable traits such as honesty, humility, and "proper behavior" (Figure \@ref(fig:cluster)). This feature might correspond, at least in part, to leaders who serve as models for their followers [see also @garfield_investigating_2019].

@@

Most potentially universal dimensions discovered here appear in a model developed by @hagen_leadership_2019, in which knowledgeable and intelligent individuals provide computational (cognitive) services to group members in exchange for a variety of fitness benefits, and these services often involve conflict resolution and other social functions [see also @garfield_evolutionary_2019]. The computational services model emphasizes mating benefits, however, which appear in `r matingbenefit`% of cultures, below our arbitrary 60% threshold for "universality" (see Figure \@ref(fig:varsupportcb)). But there was evidence for the *Social, reproductive, and materials success* feature, which included leaders in polygynous unions, having a high quality spouse, and having many children, in more cultures (`r reprosuccess`%) than any other feature (see Figure \@ref(fig:featuressupportplot)). Interestingly, the computational services model highlights women's family leadership role, and the estimated marginal mean for the *Social, reproductive, and materials success* feature was higher for females leaders than male leaders (see Figure \@ref(fig:heatmapfeaturesex)), with the caveat that there were few text records on female leaders in our sample.

@@

These factors were confounded, i.e., group context was associated with subsistence strategy, which was associated with continental region. See Figures \@ref(fig:grouptype) and S\@ref(fig:regionsubsistence). Specifically, evidence for leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013; @venkataraman_hunter-gatherer_2017]. Evidence for supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories. These patterns support perspectives suggesting residential mobility and increased costs of dispersal, subsistence intensification, and demographic transitions are associated with increasing complexity of sociopolitical organization among non-WEIRD populations [@johnson_evolution_1987; @steward_basin-plateau_1938; @powers_evolutionary_2014; @kantner_identifying_2010]. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese by @az02-001[p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in Methods section).

@@

Leaders were widely reported to receive material benefits, social status, and mating benefits but loss of material resources and social status were also widely reported, suggesting leadership to be a high-risk, high-return strategy (see Figure \@ref(fig:varsupportcb)). The results were similar for followers: followers benefited from increased material and social resources, but also often lost material and social resources. Followers were not reported to receive mating benefits, relative to leaders or other benefit types. Our results cannot determine if the benefits to leaders and followers outweighed their costs, but they do shed light on the social and biological pressures associated with leadership and social hierarchy. For both leaders and followers, individuals stood to lose what they were able to gain. See Figure \@ref(fig:varsupportcb).

@@

**Dominance and prestige**. Much of the evolutionary theoretical literature on leadership has suggested social influence is most often achieved via one of two distinct strategies: dominance or prestige [for review see @garfield_review_2019; @garfield_evolutionary_2019]. There is a long history of leadership styles being described as based on either force and strength (i.e., dominance) or respect and special skills (i.e., prestige). Such dual-models are rooted in the coupling of ethology and ethnography [see @mead_sex_1935; @sahlins_poor_1963; @lewis_leaders_1974; @tiger_dominance_1970; @bernstein_dominance_1976]. Synthesizing the ethnographic and ethological evidence, @kracke_force_1978 linked "force and persuassion" to psychological-affect between leaders and followers in small-scale societies and @barkow_prestige_1980 suggested the evolution of prestige was linked to sexual selection and symbolic culture. The dominance-prestige theory as formulated by @henrich_evolution_2001 significantly advanced evolutionary approaches to leadership and social hierarchy and inspired much subsequent work across the behavioral sciences [e.g., @van_vugt_dual_2019; @cheng_two_2013; @maner_2016; @halevy_status_2012; @suessenbach_dominance_2019; @maner_dominance_2017; @garfield_evolutionary_2019; @garfield_investigating_2019]. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on distinctions between dominance and prestige suggests, "human social stratification is characterized by the co-occurrence of both strategies." Exploratory cluster analyses reported here, however, suggest multiple features of leader qualities and leader functions, beyond dominance and prestige.

@@

<!-- DISCUSSES BOTH MANAGEMENT FEATURES -->
**Leaders are managers**. In the absence of state-institutions local leaders are often tasked with managing various economic systems and providing social services. Within the domain of leader management, there are likely to be two primary features of human leader functions: strategizing and organizing. Our results correspond very closely to what @van_vugt_cognitive_2007 define as *strategic leadership*, where leaders increase the pay-offs of another individuals behavior thereby incentivizing followership versus *coordinating leadership*, where individuals bound by group challenges must effectively organize collective effort.
